Simple Models of Metal-Line Absorption and Emission from Cool Gas Outflows

### 2010-11-17 Winds/papers/Theory/First/Submit/ms.tex
profx
git-svn-id: https://svn.ucolick.org/xsvn/trunk/radtrans@5416 ac11df5e-a788-c23b-a8e1-c758c323924d

\begin{abstract}
We analyze the absorption and emission-line profiles produced by a set
of simple, cool gas wind models motivated by galactic-scale outflow
observations.  
We implement monte carlo radiative
transfer techniques that track the propagation of scattered and
fluorescent photons to generate 1D spectra and 2D spectral images.
{\bf [Should we make the codes publicly available?  Upon request?]}
%We examine outflows with a range of power-law density and velocity
%profiles and a radiative-driven wind model.  
We focus on the \mgiid\ doublet and \ion{Fe}{2} UV1
multiplet at $\lambda \approx 2600$\AA, but the results are applicable
to other transitions that trace outflows (e.g.\ \ion{Na}{1},
\ion{H}{1} \lya, \ion{Si}{2}).   
By design, the resonance transitions show blue-shifted
absorption but one also predicts strong resonance and fine-structure line-emission at roughly
the systemic velocity.  This line-emission `fills-in' the absorption
reducing the equivalent width by up to $50\%$, shifting the centroid
by tens of \kms, and severely
reducing the effective
opacity near systemic.  Analysis of cool gas outflows that ignores this
line-emission may incorrectly infer that the gas is partially
covered, measure a significantly lower peak optical depth, and/or
conclude that gas at systemic velocity is absent   
(e.g.\ an interstellar or slowly infalling component). 
Because the \ion{Fe}{2} lines
are connected by optically-thin transitions to
fine-structure levels, their profiles
more closely reproduce the intrinsic opacity
of the wind.   Together these results naturally explain the
absorption and emission-line characteristics observed for star-forming
galaxies at $z<1$.
We also study a scenario promoted to describe the outflows
of $z \sim 3$ Lyman break galaxies and argue this model fails to
reproduce the observations when radiative transfer of scattered photons
is properly considered.   
%Strong line-emission from metal-line
%transitions is a generic prediction of wind models, even in the
%presence of dust and anisotropic configurations.  
Although line-emission complicates the analysis of absorption-line
profiles, the 
surface brightness profiles 
offer a unique means of assessing the morphology and size
of galactic-scale winds.  Furthermore, the kinematics and line-ratios
offer powerful diagnostics of outflows,  motivating deep,
spatially-extended spectroscopic observations.

@@

In a resonant line interaction, a photon excites an atom from the ground state to an excited level.  The end result is either scattering (i.e., de-excitation back to the ground state) or fluorescence (de-excitation to another excited  level).  The probability that the atom de-excites to lower level $x$ is  $p_{x} = A_{ux} / \sum_i A_{ui}$, where the $A_{ui}$ are the Einstein spontaneous emission coefficients and the sum runs over all levels accessible from the upper state $u$.  
For each interaction event, a random number was drawn to sample the final state from this probability distribution.    If the result was a scattering, the process was assumed to be coherent in the comoving frame.  If the result was fluorescence, the packet was  reemitted at the line center wavelength of the new transition.  Natural line broadening in fluorescence was ignored given the high velocity gradients in our models.  In all cases, the angular redistribution was  assumed to be isotropic.

@@

With these considerations in mind, we reanalyzed the fiducial model
with the 3D algorithm after departing from isotropy.  It is beyond the
scope of this paper to explore a full suite of anisotropic profiles.
We consider two simple examples: (i) half the fiducial model, where 
the density is set to zero for $2\pi$ steradians.  This model is
viewed from $\phi = 0^\circ$ (source uncovered) 
to $\phi = 180^\circ$ (source covered); and 
(ii) a bi-conical wind where the density is set to zero for angles
$|\theta| < \theta_b$ and that is viewed along the axis of rotational
symmetry ($\theta = 90^\circ$ is defined to be along our sightline).

@@

We have also analyzed the predicted profiles for a series of biconical
winds with $\theta = 10-60^\circ$, each viewed along the axis of
rotational asymmetry with the wind covering the
source. Figure~\ref{fig:biconical} and
Table~\ref{tab:line_diag} summarizes the results for several cases.  For
$\theta_b = 10^\circ$, the results are
very similar to the fiducial model;  the profiles show
similar equivalent widths for absorption and emission.  The key
quantitative difference is that the \feiis\ line-emission is
suppressed at $\delta v \approx 0 \mkms$ because the portion of the
wind with that projected velocity has been removed.  For
$\theta_b>45^\circ$, enough of the wind has been removed that the
line-emission is several times weaker than the fiducial model.
Similarly, the absorption-line profiles more closely track the
intrinsic optical depth of the outflow.   Of course, these same models
when viewed `sideways' would show no absorption yet significant
line-emission.  Such events are sufficiently rare (Rubin et
al., in prep) that we consider this level of anisotropy to be
uncommon.

@@

Another factor that may reduce line-emission is %an
%anisotropic 
wind %model
anisotropy. The flux is lower, for example, if one
eliminates the backside to the wind (e.g.\ the source itself could
shadow a significant fraction of the backside).  
Similarly, a biconical
wind can have significantly lower line-emission for the fiducial model
($\S$~\ref{sec:anisotropic}). 
For the anisotropic winds explored in this paper,
the line-emission scales 
roughly as $\Omega/4\pi$ where $\Omega$ is the angular extent of the
wind\footnote{There is also a first order dependence on the actual
  orientation of the wind, with maximal emission when the backside is
  fully viewed.}.  Because we require that the wind 
points toward us, it is difficult to reduce $\Omega$ much
below $2 \pi$ and, therefore, anisotropy reduces the
emission by a factor of order unity.  
Similarly, an anisotropically emitting source would modify the predicted
line-emission.  For example, if the backside of the galaxy were
brighter/fainter then would one predict brighter/fainter emission
relative to the observed continuum.
To reduce the emission, the brightest regions
of the galaxy would need to be oriented toward Earth.  Although this
would not be a
generic orientation, most spectroscopic samples are magnitude-limited
and biased towards detecting galaxies when observed at their
brightest.  In principle, this might imply a further
reduction of order unity.

@@

Figure~\ref{fig:obs_ew} demonstrates one observational
consequence: reduced measurements for the absorption
equivalent width \ewabs\ of the flow.  In the case of \ion{Mg}{2},
which has the most strongly affected transitions, \ewabs\ is reduced by
$30-50\%$ from the intrinsic equivalent width.  In turn, one may derive 
a systematically lower optical depth or velocity extent for the wind,
and therefore a lower total mass and kinetic energy.  
The effects are most pronounced for wind
scenarios where the peak optical depth occurs near $\mdv = 0 \mkms$.
Geometric projection limits the majority of scattered photon emission to
have $|\mdv| < 200 \mkms$; therefore,  the absorption profiles
are filled-in primarily at these velocities.

@@

\item The line-emission occurs preferentially at the systemic velocity
  of the galaxy and `fills-in' the absorption profiles at velocity
  offsets $|\mdv| < 200 \mkms$ from systemic.  For transitions that
  are only coupled to the ground-state (e.g.\ \ion{Mg}{2}, \lya,
  \ion{Na}{1}),  this implies much lower, absorption-line equivalent widths (by up to
  $50\%$) and also velocity profiles that are significantly offset
  from the intrinsic optical depth profile.

@@

\item We examined two scenarios designed to mimic the wind model
  proposed by \cite{steidel10} for $z \sim 3$ Lyman break galaxies.  Our
  implementation of this model genreates substantial line-emission
  from scattered photons that greatly modifies the predicted
  line-profiles so that this model does not reproduce the observed
  line-profiles.

@@

\begin{figure}
%\epsscale{0.8}
%\plotone{../Figures/fig_asymm_spec.ps}
\includegraphics[width=5.5in]{f9.eps}
\caption{
Profiles of the \ion{Fe}{2} and \ion{Mg}{2} profiles for the fiducial
case (black lines) compared against a biconical wind with varying
opening angle ($\theta_b = 0^\circ$ is a full wind).  For modest angles, one
observes similar results to the fiducial model but larger angles
($\theta_b > 45^\circ$) the emission is significantly suppressed at
all velocities.  
}
\label{fig:biconical}
\end{figure}

@@

\begin{figure}
%\includegraphics[scale=0.6,angle=90]{../Figures/fig_obs_ew.ps}
\includegraphics[scale=0.6,angle=90]{f21.eps}
\caption{
Difference between the `observed' absorption equivalent width \ewabs\
which includes the flux of scattered photons and the `intrinsic'
equivalent width $W_{\rm i}$ that ignores photon scattering.  The
dashed (dotted) curves trace a 50\%\ (10\%) reduction in \ewabs\
relative to $W_{\rm i}$.  One notes a reduction in \ewabs\ by $\approx
30-50\%$ for the \mgiia\ transition (the effect is generally larger
for \mgiib).  The effects of scattered photons are reduced for the
\ion{Fe}{2} transitions because a fraction (in fact a majority for
\feiia) of the absorbed photons fluoresce as \feiis\ emission at
longer wavelengths and do not `fill-in' the absorption profiles.
}
\label{fig:obs_ew}
\end{figure}

>>>>

\begin{abstract}
We analyze the absorption and emission-line profiles produced by a set
of simple, cool gas wind models motivated by galactic-scale outflow
observations.  
We implement monte carlo radiative
transfer techniques that track the propagation of scattered and
fluorescent photons to generate 1D spectra and 2D spectral images.
%{\bf [Should we make the codes publicly available?  Upon request?]}
%We examine outflows with a range of power-law density and velocity
%profiles and a radiative-driven wind model.  
We focus on the \mgiid\ doublet and \ion{Fe}{2} UV1
multiplet at $\lambda \approx 2600$\AA, but the results are applicable
to other transitions that trace outflows (e.g.\ \ion{Na}{1},
\ion{H}{1} \lya, \ion{Si}{2}).   
By design, the resonance transitions show blue-shifted
absorption but one also predicts strong resonance and fine-structure line-emission at roughly
the systemic velocity.  This line-emission `fills-in' the absorption
reducing the equivalent width by up to $50\%$, shifting the centroid
by tens of \kms, and severely
reducing the effective
opacity near systemic.  Analysis of cool gas outflows that ignores this
line-emission may incorrectly infer that the gas is partially
covered, measure a significantly lower peak optical depth, and/or
conclude that gas at systemic velocity is absent   
(e.g.\ an interstellar or slowly infalling component). 
Because the \ion{Fe}{2} lines
are connected by optically-thin transitions to
fine-structure levels, their profiles
more closely reproduce the intrinsic opacity
of the wind.   Together these results naturally explain the
absorption and emission-line characteristics observed for star-forming
galaxies at $z<1$.
We also study a scenario promoted to describe the outflows
of $z \sim 3$ Lyman break galaxies and argue this model fails to
reproduce the observations when radiative transfer of scattered photons
is properly considered.   
%Strong line-emission from metal-line
%transitions is a generic prediction of wind models, even in the
%presence of dust and anisotropic configurations.  
Although line-emission complicates the analysis of absorption-line
profiles, the 
surface brightness profiles 
offer a unique means of assessing the morphology and size
of galactic-scale winds.  Furthermore, the kinematics and line-ratios
offer powerful diagnostics of outflows,  motivating deep,
spatially-extended spectroscopic observations.

@@

In a resonant line interaction, a photon excites an atom from the
ground state to an excited level.  The end result is either scattering
(i.e., de-excitation back to the ground state) or fluorescence
(de-excitation to another excited  level).  The probability that the
atom de-excites to lower level $x$ is  $p_{ux} = A_{ux} / \sum_i A_{ui}$, where the $A_{ui}$ are the Einstein spontaneous emission coefficients and the sum runs over all levels accessible from the upper state $u$.  
For each interaction event, a random number was drawn to sample the
final state from this probability distribution.    If the result was a
scattering, the process was assumed to be coherent in the comoving
frame.  If the result was fluorescence, the packet was  reemitted at
the line center wavelength of the new transition.  Natural line
broadening in fluorescence was ignored given the high velocity
gradients in our models.  In all cases, the angular redistribution was
assumed to be isotropic.

@@

With these considerations in mind, we reanalyzed the fiducial model
with the 3D algorithm after departing from isotropy.  It is beyond the
scope of this paper to explore a full suite of anisotropic profiles.
We consider two simple examples: (i) half the fiducial model, where 
the density is set to zero for $2\pi$ steradians.  This model is
viewed from $\phi = 0^\circ$ (source uncovered) 
to $\phi = 180^\circ$ (source covered); and 
(ii) a bi-conical wind which fills 
$|\theta| < \theta_b$ into and out of the plane of the sky and that is
viewed along the axis of rotational symmetry ($\theta = 0^\circ$ is
defined to be along this axis)..

@@

We have also analyzed the predicted profiles for a series of bi-conical
winds with $\theta_b = 30-80^\circ$, each viewed along the axis of
rotational asymmetry with the wind covering the
source. Figure~\ref{fig:biconical} and
Table~\ref{tab:line_diag} summarizes the results for several cases.  For
$\theta_b = 80^\circ$, the results are
very similar to the fiducial model;  the profiles show
similar equivalent widths for absorption and emission.  The key
quantitative difference is that the \feiis\ line-emission is
suppressed at $\delta v \approx 0 \mkms$ because the portion of the
wind with that projected velocity has been removed.  For
more highly collimated winds
$\theta_b<45^\circ$, however, the
line-emission is several times weaker than the fiducial model.
Similarly, the absorption-line profiles more closely track the
intrinsic optical depth of the outflow.   Of course, these same models
when viewed `sideways' would show no absorption but strong
line-emission.  Such events are sufficiently rare (Rubin et
al., in prep) that we consider this level of anisotropy to be
uncommon.

@@

Another factor that may reduce line-emission is %an
%anisotropic 
wind %model
anisotropy. The flux is lower, for example, if one
eliminates the backside to the wind (e.g.\ the source itself could
shadow a significant fraction of the backside).  
Similarly, a bi-conical
wind can have significantly lower line-emission for the fiducial model
($\S$~\ref{sec:anisotropic}). 
For the anisotropic winds explored in this paper,
the line-emission scales 
roughly as $\Omega/4\pi$ where $\Omega$ is the angular extent of the
wind\footnote{There is also a first order dependence on the actual
  orientation of the wind, with maximal emission when the backside is
  fully viewed.}.  Because we require that the wind 
points toward us, it is difficult to reduce $\Omega$ much
below $2 \pi$ and, therefore, anisotropy reduces the
emission by a factor of order unity.  
Similarly, an anisotropically emitting source would modify the predicted
line-emission.  For example, if the backside of the galaxy were
brighter/fainter then would one predict brighter/fainter emission
relative to the observed continuum.
To reduce the emission, the brightest regions
of the galaxy would need to be oriented toward Earth.  Although this
would not be a
generic orientation, most spectroscopic samples are magnitude-limited
and biased towards detecting galaxies when observed at their
brightest.  In principle, this might imply a further
reduction of order unity.

@@

Figure~\ref{fig:obs_ew} demonstrates one observational
consequence: reduced measurements for the absorption
equivalent width \ewabs\ of the flow.  In the case of \ion{Mg}{2},
which has the most strongly affected transitions, \ewabs\ is reduced by
$30-50\%$ from the intrinsic equivalent width (the equivalent width
one would measure in the absence of scattered photons).  In turn, one may derive 
a systematically lower optical depth or velocity extent for the wind,
and therefore a lower total mass and kinetic energy.  
The effects are most pronounced for wind
scenarios where the peak optical depth occurs near $\mdv = 0 \mkms$.
Geometric projection limits the majority of scattered photon emission to
have $|\mdv| < 200 \mkms$; therefore,  the absorption profiles
are filled-in primarily at these velocities.

@@

\item The line-emission occurs preferentially at the systemic velocity
  of the galaxy and `fills-in' the absorption profiles at velocity
  offsets $|\mdv| < 200 \mkms$ from systemic.  For transitions that
  are only coupled to the ground-state (e.g.\ \ion{Mg}{2}, \lya,
  \ion{Na}{1}),  this implies much lower, absorption-line equivalent widths (by up to
  $50\%$) and observed absorption profiles that are significantly offset in velocity
  from the intrinsic optical depth profile.

@@

\item We examined two scenarios designed to mimic the wind model
  proposed by \cite{steidel+10} for $z \sim 3$ Lyman break galaxies.  Our
  implementation of this model genreates substantial line-emission
  from scattered photons that greatly modifies the predicted
  line-profiles so that this model does not reproduce the observed
  line-profiles.

@@

\begin{figure}
%\epsscale{0.8}
%\plotone{../Figures/fig_asymm_spec.ps}
\includegraphics[width=5.5in]{f9.eps}
\caption{
Profiles of the \ion{Fe}{2} and \ion{Mg}{2} profiles for the fiducial
case (black lines) compared against a bi-conical wind with varying
opening angle ($\theta_b = 0^\circ$ is a full wind).  For modest angles, one
observes similar results to the fiducial model but larger angles
($\theta_b > 45^\circ$) the emission is significantly suppressed at
all velocities.  
}
\label{fig:biconical}
\end{figure}

@@

\begin{figure}
%\includegraphics[scale=0.6,angle=90]{../Figures/fig_obs_ew.ps}
\includegraphics[scale=0.6,angle=90]{f21.eps}
\caption{
Difference between the `observed' absorption equivalent width \ewabs\
which includes the flux of scattered photons and the `intrinsic'
equivalent width $W_{\rm i}$ that ignores photon scattering.  The
dashed (dotted) curves trace a 50\%\ (10\%) reduction in \ewabs\
relative to $W_{\rm i}$.  One notes a reduction in \ewabs\ by $\approx
30-50\%$ for the \mgiia\ transition (the effect is generally larger
for \mgiib).  The effects of scattered photons are reduced for the
\ion{Fe}{2} transitions because a fraction (in fact a majority for
\feiia) of the absorbed photons fluoresce as \feiis\ emission at
longer wavelengths and do not `fill-in' the absorption profiles.
The figure shows results for all of the models presented in
Tables~\ref{tab:line_diag} and \ref{tab:plaw_diag}.
}
\label{fig:obs_ew}
\end{figure}
